\documentclass[11pt, a4paper]{article}

\input{common}

\usepackage[left=2cm, right=2cm, top=2cm, bottom=2cm]{geometry}
\usepackage{enumitem}

\addbibresource{references.bib}

\docname{ethics-application-working-draft}

\author{James Thompson\\Te Herenga Waka --- Victoria University of Wellington}
\date{\today}

\begin{document}

\title{Ethics Application: Aligning LLMs to NZ Public Values}
\maketitle

\abstract{
	Simple document to just help me write down the answers to the questions that are in the ethics application form. This is not the final version of the ethics application, but a working draft to help me organise my thoughts and ideas.
}

\section{Documents included}

The final submission will include two additional documents: the participant information sheet and a sample interview script. The participant information sheet is included in 

\section{Project details}

\subsection{Q9. Describe the aims and objectives of this project.}

\subsection{Q10. Describe the background, benefits and scholarly value of the project}

\subsection{Q11. Explain any ethical issues your research raises for participants, yourself as the researcher, or wider communities and institutions, and how you will address these}

\section{data collection and recruitment}

\subsection{Q16. Provide an explanation of the sampling rationale for your study}

\subsection{Q17. How many participants will be involved in your research?}

\subsection{Q18. What are the characteristics of the people you will be recruiting?}

\subsection{Q19. Outline in detail the method(s) of recruitment you will use for participants in your study.}

\subsection{Q20. Explain the details of the method of data collection.}

\section{Te Tiriti o Waitangi Statute}

\subsection{Q28. Describe how your research will uphold our Te Tiriti o Waitangi Statute, in practice.}

\section{Project risks}

There is a chance this project might use deception so that the participants won't know it is AI.

A risk of knowing participants if there will be volunteers who are recruited (i.e friends and family).

\end{document}
